Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-05 WSSSPE1_report/WSSSPE1_summary_report.tex
Remaining paper summaries for Implementing Policy.

\begin{abstract}
This document discusses the First Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE1).  The content
is based on the workshop itself, as recorded in a set of
collaborative notes taken during the workshop, including discussion
about the keynote presentations, three themed panels that focus on
papers accepted by the workshop, other discussion, cross-cutting
issues, use cases, and conclusions.  It joins a nascent literature
seeking to understand what drives software work in science and how the
reward systems of science thereby shapes the type of software work
undertaken, including the extent to which developers are motivated to
build software for the long-term, for the use of others, and whether
to work collaboratively or separately. Unique perspectives were
captured about issues such as documentation of APIs, software
deployment, and development practices; software licenses; career
paths for scientific software developers; identifiers for software
through a mechanism such as a Digital Object Identifier or
publication of a ``software paper''; and evidence of software
contribution and impact, recorded in online systems such as source
code repositories like GitHub. For truly sustainable software there
should be no endpoint, as the software products continue to be used
and useful beyond the single institution, grant, and developer or
development team.

@@

%\note{papers and presentations and URLs should be references - see
%  figshare-web in .bib file for URL style example}

@@

Additional volunteers:
\begin{itemize}
% \item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
\item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
\item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
\item * Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
\item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
% \item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
\item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
\item * Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
\item * Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
% \item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
\item * Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
% \item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
\item * Frank L\"{o}ffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
\item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
\item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
\item * James Howison $\langle$\url{james@howison.name}$\rangle$
\item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
\item * Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
\item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
\item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
\item * Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
\end{itemize}

@@

%\pagebreak
%
%\section*{Executive Summary}
%
%\todo{1 page summary goes here}
%
%\pagebreak

@@

Because progress in scientific research is dependent on the quality of and
accessibility to software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and debates have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
document. An estimate of 90 to 150 participants were present
throughout the day. Additional papers based on extended versions of workshop
submissions are expected. The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC14 conferences), to obtain
funds from the National Science Foundation and the Gordon and Betty Moore
Foundation to support these workshops, and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress. Additionally, a minisymposium that aims to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the 2014 Society for Industrial and Applied Mathematics
(SIAM) Annual Meeting on ``Reliable
Computational Science.''

The goal of this paper is to summarize the contents of the workshop papers
and discussion, and identify issues that future WSSSPE events will address.
Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda. Section~\ref{sec:process} is
based on that publication. The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations (\S\ref{sec:keynotes}), the three
panels (\S\ref{sec:devel}-\ref{sec:community}), 
cross-cutting issues (\S\ref{sec:cross-cutting}),
use cases (\S\ref{sec:use-cases}), and conclusions
(\S\ref{sec:conclusions}).

@@

\begin{quote}
``In practice, scientific software activities are part of an ecosystem
where key roles are held by developers, users, and funders. All three
groups supply resources to the ecosystem, as well as requirements that
bound it. Roughly following the example of NSF's Vision and Strategy
for Software~\cite{NSF_software_vision},
the ecosystem may be viewed as having challenges related to:

\begin{itemize}[leftmargin=0.2in]
\item the development process that leads to new (versions of) software
\begin{itemize}[leftmargin=0.2in]
\item how fundamental research in computer science or
  science/engineering domains is turned into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of
    scientific software and vice versa
\end{itemize}
\item the support and maintenance of existing software, including
  software engineering
\begin{itemize}[leftmargin=0.2in]
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and
  sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}[leftmargin=0.2in]
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}[leftmargin=0.2in]
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and multiple countries,
  such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of
  publishers
\end{itemize}
\item education and training''
\end{itemize}

@@

\begin{quote}
``We invite short (4-page) position/experience reports that will be used
to organize panel and discussion sessions. These papers will be
archived by a third-party service, and provided DOIs [Digital Object Identifiers].
We encourage
submitters to license their papers under a Creative Commons license
that encourages sharing and remixing, as we will combine ideas (with
attribution) into the outcomes of the workshop.  An interactive site
will be created to link these papers and the workshop discussion, with
options for later comments and contributions. Contributions will be
peer-reviewed for relevance and originality before the links are added
to the workshop site; contributions will also be used to determine
discussion topics and panelists. We will also plan one or more papers
to be collaboratively developed by the contributors, based on the
panels and discussions.''
\end{quote}

@@

The organizers decided to list 54 of the papers as significant
contributions to the workshop, a very high acceptance rate but one
that is reasonable, given the goal of broad participation and the fact
that the reports were already self-published. The papers were
grouped into three main categories, namely \emph{Developing and
Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
subject area was associated with a panel and discussion at the
workshop.

@@

@@

Communities are extremely important in software projects, and both their
building and continued engagement need attention during the project life cycle.
Several of the submitted papers discussed how communities have been built around
projects, and what is needed to enable a project to grow. Among these are public
source code hosting, mailing lists, documentation, wikis, bug trackers, software
downloads, continuous integration, software quality dashboards, and of course,
a general web presence to tie all of these things together. Questions were posed
around users not answering the questions of other users. Several participants
offered counterexamples, such as mailing lists where developers do not participate
as much due to users being more active, or whether the ``core
team'' can end up setting unrealistic expectations by doing too much. Team Geek~\cite{opac-b1134063}
and Turk's paper on scaling code in the human dimension~\cite{Turk:2013:SCH:2484762.2484782}  discuss how development
lists that are welcoming to people actually have many more people contributing.

@@

\todo{some text is needed here before the first subsection}

\subsection{Modeling Sustainability}

%\todo{seems like definitions will be mentioned elsewhere in the paper?}
Each of the framework papers included a definition of sustainability, with many overlaps.  In essence the papers related the question to change over time, an idea well captured in the definition used by the UK's Scientific Software Sustainability Institute: ``software you use today will be available---and continue to be improved and supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}. This idea was also expressed as software that continues to serve its users~\cite{Pierce_WSSSPE}.
See \S\ref{sec:cross-cutting} for more on the discussion of definitions of sustainability in the workshop.

\subsection{Credit, Citation, Impact}

\subsection{Implementing Policy}

Two of the papers in this group focus on specific facets of software
design that are important factors in a project's sustainability but
are often addressed only late in the scientific software development
cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API
governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity
models for software security. The other two papers discuss
implementation strategies for science from the perspective of
facilitating many or all facets of sustainanability-oriented software
design for science: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
large projects that have software infrastructue development built-in
with cross-training domain scientist PIs in software engineering best
practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
specialized software engineering skills could be turned into shared
instrumentation with low barriers to access.

% Finally, we’ll look at some papers that discuss implementing sustainability in scientific software.

% Blanton and Lenhardt from the Renaissance Computing Institute discuss these issues from a user perspective.

% The authors focus on a point that has been brought up many times in this context:
% There is a tension between writing code that is good enough just to “get it done,” i.e. to publish a paper about scientific results obtained using software, 
% and “getting it right,” that is, developing software that is comprehensible to future users and reviewers.
% Just because the elevator panel works like this doesn’t mean it’s sustainable for the long run.
% We don’t have a way to validate that the software used in a paper is actually done right.
% The best way to get software designed correctly is to make sure best practices are considered from the start.

% The authors highlight two models for sustainable software, at different extremes:
% One is what they call “co-funding”:
% In these projects, usually large, multi-year collaborations, 
% there is equal emphasis on both the science and the software development. 
% Both are planned into the project from inception. 
% In the life sciences, the iPlant Collaborative, Galaxy Project, and Qiime are good examples of these sorts of large, well-designed projects.

% At the other extreme, they discuss “software carpentry”: 
% in this model, it’s assumed that the scientists themselves will write and maintain their code.
% Groups like Software Carpentry and ROpenSci assume that 
% scientists won’t have access to dedicated software engineering, 
% so they try to give them tools to use best practices in their own software development.

% There might be a middle ground here: a way to get the engineering expertise that large co-funded projects have to individual scientist-developers.
% Hilmar Lapp of NESCent and I discuss one such possibility in our paper, Software Engineering as Instrumentation for the Long Tail of Scientific Software.

% What do we mean when we refer to the “long tail” of scientific software?
% Think of the distribution of resources in scientific software. Most are focused on big projects with lots of community buy-in and funding. But a lot of scientific software exists away from this model.
% For example, scientific software can be used long after the original developer has moved on or the funding runs out.
% Look at MacClade: it was originally released in 1986 and last updated in 2005,
% but it was still cited over 400 times in 2013!
% The scientists who developed it have a newer package, Mesquite, that was meant to replace MacClade, but they haven’t had sufficient time or resources to maintain either package fully, let alone both of them.

% Another dimension of the long tail can also be found in my particular research domain. 
% In the field of phylogenetics, we have a lot of programs that implement different computational methods in slightly different ways.
% Here, Joe Felsenstein has listed some (but not anywhere near all) phylogenetics packages available online.
% Most of these programs are developed by academic scientists…
% They generally have limited training in software engineering
% Limited time or career incentive to improve software
% Limited funding

% So, to summarize a bit:
% Making sustainable software means we have to pay attention to many facets of software design, like APIs, security, user experience, testing, etc.
% A single project that requires one full­-time software engineer may actually require fractions of different kinds of engineers. 
% But long-tail projects can’t even fund one FTE, let alone one that can address all these facets.

% Then we have to consider that the users of scientific software are scientists, 
% so the developers need to understand the users and the science.
% This is the idea of a “t-skilled” person: 
% one who is both well-versed in a scientific domain 
% and deeply experienced in one or more facets of software engineering. 
% These people are pretty rare in the first place and difficult to retain in academia, because the academic career structure doesn’t incentivize this.

% We should look at software engineering as an expensive resource, but one that needs to be accessible to scientists at all levels.
% Think of it as analogous to DNA sequencing:
% Sequencers used to be something that individual labs and institutions had to 
% buy, maintain, and operate themselves, 
% so only highly-funded operations had them and probably didn’t use them to their full capacity even when they had one.
% But now, core facilities provide the instrumentation and service to labs of any size. 
% Anyone can pay a core facility to sequence their samples for them and provide quality control and bioinformatics advice as additional services.

% We propose that software engineering can be “instrumented” in a similar way. 
% Let’s create a nonprofit center for scientific software engineering.
% This center can hire these t-skilled personnel and provide access to them for projects at contracted cost.
% Because the center is focused on providing development services to scientific projects, it is not tied to the long-term success or failure of any individual project.
% It would emphasize the centrality of doing good science by making functional software tools as envisioned by scientists.

% So, to conclude…
% Implementing policies to encourage sustainability in scientific software 
% requires that many facets of good software design are addressed throughout the lifecycle of these projects. 
% But most of them aren’t addressed in the status quo.
% We’ve highlighted some of these facets today and suggested some possible solutions.
% Large projects can afford to hire software engineers with the expertise to implement these facets correctly.
% Grassroots developer groups can provide guidance to scientists about best practices in software development.
% We think there is a place for a software engineering center that can provide 
% both engineering expertise and guidance 
% with a contract-driven instrumentation model 
% to the scientific software in the long tail.

%There were a number of papers categorized under the Communities banner and so
%two presenters each summarized half of the submissions in this area.
\todo{an intro to the whole section is needed}

\subsection{Communities}

%This collection of papers was summarized by Karen Cranston.

%\subsection{Communities Part 2}

%Nancy Wilkins-Diehr summarized 6 papers in this part.

\subsubsection{What are communities?}

\subsubsection{Challenges of community}
include metrics for community success (e.g., more external
contributors than internal).

\subsubsection{Admirable scientific software communities}
\begin{itemize}
\item example communities
\item their (quick) origin stories.
\end{itemize}

\subsubsection{Resources for learning about software communities}
\begin{itemize}
\item academic fields studying communities (and software communities).
\item courses on online communities
\item need for software carpentry module on organizing communities?
\item \cite{howison_scientific_2011}
\item \cite{citeulike:11831265}
\item \cite{citeulike:7888211}
\item \cite{citeulike:478633}
\item \cite{opac-b1134063}
\item \cite{DBLP:conf/cscw/HowisonH11}
\item \cite{DBLP:conf/cscw/2011}
\item \cite{Trapani:2011}
\item \cite{Allsopp:2012}
\item \cite{1749-4699-6-1-015010}
\item \cite{Turk:2013:SCH:2484762.2484782}
\end{itemize}

\subsection{Industry \& Economic Models}

\todo{text about papers and discussion in this part of the workshop}

\subsection{Education \& Training}

\todo{text about papers and discussion in this part of the workshop}

In this subtopic, two papers included discussion of the definition of
sustainability.  Venters et al.~\cite{Venters_WSSSPE} mentioned that this is a
rather ambiguous concept, and that the lack of an accepted definition gets in
the way of integrating the concept into software engineering. They suggest that
sustainability is a non-functional requirement, and that the quality of
software architectures determines sustainability. They then propose that
sustainability is a consequence of a set of central quality attributes:
extensibility, interoperability, maintainability, portability, reusability, and
scalability. Finally, they develop an architecture evaluation framework based
on scenarios that help to illuminate how to measure quality or sustainability
at the architectural level.

In this section, we discuss some of the software projects as case studies to
better understand the points discussed during the workshop and described in the
previous sections. We dive deeper to find how are they affected by the
sustainability related issues in practice. Additionally, we discuss issues
surrounding people, especially the career paths of original developers and
investigate to answer the question: are the original developers are still
around?

We classify the software projects discussed in the workshop in two broad
categories. First, the \emph{utility} software, comprising of general purpose
software. Utility software is often used as enabler and/or facilitator for the
development of other tools and techniques to carry out scientific work. This
includes the software developed to efficiently utilize new research
infrastructures. Second, the \emph{scientific} software, comprising the
software that was originally developed with an aim to solve a specific
scientific problem.

% Why are we classifying software like above?
This classification is motivated by our argument that the two kinds of software
projects wildly vary in factors such as scope, purpose and usage. The
development and management of each kind is significantly different.
Consequently, the sustainability challenges faced by them differ and must be
treated separately. For instance, the challenges faced by a gateway software
development project such as \textit{CIPRES} or a visualization software product
such as \textit{VisIT} are distinct to a niche software for \textit{ab initio}
modeling and simulation such as \textit{VASP} or \textit{Quantum Espresso}.

\subsection{Utility Software}
Software developed with a potentially wider audience and general purpose usage
in mind is utility software. Utility software typically does not address
fundamental research problems for a given scientific domain. Examples are
collaborative development frameworks such as \emph{GitHub} and
\emph{Bitbucket}, distributed workflow and generic computing frameworks
such as \emph{Galaxy}, \emph{nanoHUB}, \emph{SimGrid}, \emph{Swift},
\emph{Globus} and \emph{VisTrails}, and visualization frameworks such as
\emph{VisIT}.

% What are the sustainability challenges? What are the oppurtunities?
Development is often a high risk/reward undertaking exploring uncharted
territories and is largely influenced by (re)usability factors. Owing to
relatively large number of features, the development and prototype process is
also lengthy which poses a significant survival risk. Challenges on a class of
utility software for new architectures is well discussed
in~\cite{Ferenbaugh_WSSSPE}.

On the other hand, utility software presents oppurtunities to be usable by a
larger community making its undertaking and development an attractive pursuit.
It is generally more visible in community which in turn leads to a broader and
deeper participation. For instance, promoting collaborations across the breadth
(e.g.  different departments) and depth (e.g. stakeholders within a department)
of community, one of the key ingredient of a sustainable process. Successful
utility projects reap high rewards and have a longer usage span. Development
process becomes user-driven and self-sustaining.

One such example is the \emph{Galaxy} project. It follows agile software
development practices and implements standards such as test-driven development
and socialized bug managing practices via \emph{trello}. Galaxy
\emph{histories} and \emph{toolshed} offers easy community sharing of data and
tools further promoting a collaborative environment. The project very closely
follows the guidelines described in~\cite{Carver_WSSSPE} and many
from~\cite{Prlic_WSSSPE}. Many utility software projects are often developed
aiming better utilizing a particular, new infrastructure and architecture,
e.g., \emph{MVAPICH}, \emph{VisIT}. Similarly, to leverage the power of
emerging architectures such as accelerators, new code and libraries are
required. The experience of one such effort as described
in~\cite{Ferenbaugh_WSSSPE} which met with a limited success but nonetheless
with many invaluable lessons were learned about influential cultural and
technical aspects in sustainable software development practices.

A relatively new paradigm in utility software is the software delivered as
service over the web. With increasing popularity of cloud-based storage and
computational environments, many users are leaning towards tools used as
services. \emph{GitHub} and \emph{Bitbucket} can be argued to be
such tools, catering to collaborative development. For scientific users
\emph{Globus}-based tools are a case of service-based utility software
discussed during the workshop. The data movement and sharing services offered
by Globus can be easily used over the web by collaborating researchers.
%
%
%\note{Dan: so is the only difference here that this category is aimed at new
%systems, where utility software is not?  If so, perhaps this should be
%incorporated into that subsection, so there would just be two subsections.
%Something in the framing of this subsection doesn't feel right to me.  FL: I
%respectfully disagree the the only reason to develop frameworks is to better
%utilize (new or old) infrastructure. Another really important reason is to
%modularize development, making it much easier for an individual developer/user
%to integrate their own code. Ketan: I think in this section I wanted to bring
%up a class of software projects which are undertaken with the main aim of
%tackling new systems and architectures in mind. New infrastructures such as
%GPU, Cloud, fast IB, federal science networks, etc. were on my mind as I see
%many submissions around them in the workshop.}
%
%

\subsection{Scientific software}
Scientific software consists primarily of special-purpose software that was
purpose-built for a target use-case scenario,  fixed/frozen requirements in
mind or solving a specific problem. Software projects pertaining to specific
scientific domain tend to be in a niche and the user community tends to be
small to medium. They are mostly driven by the science and specific needs of a
research group. Specific needs such as numerical accuracy and
algorithmic optimization are some of the paramount requirements of most
scientific software.

% What are the challenges?
Long-term sustainability of scientific software is often a significant
challenge and face radically different issues compared to utility software.
Many submissions reported that software is practically considered a `byproduct'
of the actual research. Others contended that the software was not the main
funded part of their research. A smaller codebase and fixed requirements result
in stability, ease of installation, and configurability.  Many such projects
mature and are treated as libraries to be used into larger systems such as some
of the utility software discussed in the previous section. While the software
can stay stable and require relatively low maintenance, the responsibility is
often on the shoulders of a very few developers who might not be specialists in
software development. Development tends to be linear and simplistic with a
limited scope to follow software best practices.

Some examples of such software discussed as part of the workshop are,
\emph{DUNE}, \emph{R/qtl}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP},
\emph{FASTMath}.  Most of the above efforts are focused on one scientific
and/or applied mathematics domain. However, sometimes such projects grow beyond
the initial vision of developers. One such example is \emph{Kitware}, which
while being a software product specializing in the scientific process, has a
core focus of developing communities around software processes. An instance of
this process is the development of the \emph{CMake} build utility, which
started out as a building tool for \emph{ITK} but grew to become a generic
build utility for C++ projects. Similarly, \emph{PETSc} is growing towards
becoming a general purpose utility system usable for solving variety of
scientific problems.
%

Finally, it appears that the career options of developers from different kinds
of software projects differ significantly. In order to further investigate the
much discussed issue of career paths for developers of these software, we
conducted a small survey of the use-cases discussed above with respect to the
current status of its core developers. We looked at the workshop submissions
and the original cited papers for the respective software. Where available, we
looked at the webpages and source code of the earliest versions to find the
original developers. We included projects older than 5 years and excluded
relatively new projects such as \emph{SciPy} and \emph{numFOCUS} from this
survey. we find that in a majority of scientific software more than one
original developers are still around, in some cases for more than 10 years,
(eg. \emph{R/qtl}).  Whereas, in the case of utility software projects, it
appears that at least one original developer is still around, often in a
part-time or ``benefactor" role. In other cases, it appears that the original
developers have went on to assume a larger role, with the same project.

In conclusion, we find that there are distinctions in the characteristics and
the kinds of challenges faced between utility and scientific software projects.
In terms of career paths of original developers, it appears that the career
choices for utility software developers are broader compared to the developers
of scientific software. However, it appears that in the case of scientific
software, core competencies of original developers tend to be domain science
who happen to be the coders of their software. As projects as a whole, the
small-scale, specific software have a lower risk and cost in terms of
sustaining as compared to utility software. We also find that the utility
software tend to adhere more loyally to the many best practices discussed
during the workshop. Often the sustainability of scientific software projects
is achieved by the fact that the core developer or team heavily utilizes the
software for their science, e.g.,  \emph{R/qtl, PETSc}.  Furthermore, the
development of scientific software requires more scientific skills compared
with those of utility software. This diminishes the need of specialist
programmers and practices in many cases where the bulk of development is
carried out by a domain scientist. With these inferences, we believe that
separate guidelines and/or principles for sustainability be defined and
practiced for each category of scientific software and they be given separate
treatment.

\subsection{Top Issues}

\appendix
\section{Papers Accepted and Discussed at WSSSPE1}

\section*{Developing and Supporting Software}

@@

\section*{Policy}

>>>>

\begin{abstract}
This document discusses the First Workshop on Sustainable
Software for Science: Practice and Experiences (WSSSPE1).  The content
is based on the workshop itself, as recorded in a set of
collaborative notes taken during the workshop, including discussion
about the keynote presentations, three themed panels that focus on
papers accepted by the workshop, other discussion, cross-cutting
issues, use cases, and conclusions.  It joins a nascent literature
seeking to understand what drives software work in science and how the
reward systems of science thereby shape the type of software work
undertaken, including the extent to which developers are motivated to
build software for the long-term, for the use of others, and whether
to work collaboratively or separately. Unique perspectives were
captured about issues such as documentation of APIs, software
deployment, and development practices; software licenses; career
paths for scientific software developers; identifiers for software
through a mechanism such as a Digital Object Identifier or
publication of a ``software paper''; and evidence of software
contribution and impact, recorded in online systems such as source
code repositories like GitHub. For truly sustainable software there
should be no endpoint, as the software products continue to be used
and useful beyond the single institution, grant, and developer or
development team.

@@

\note{papers and presentations and URLs should be references - see
  figshare-web in .bib file for URL style example}

@@

Additional volunteers:
\begin{itemize}
\item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
\item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
\item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
\item Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
\item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
\item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
\item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
\item * Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
\item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
\item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
\item * Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
\item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
\item * Frank L\"{o}ffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
\item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
\item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
\item * James Howison $\langle$\url{james@howison.name}$\rangle$
\item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
\item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
\item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
\item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
\item * Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
\end{itemize}

@@

\pagebreak

\section*{Executive Summary}

\todo{1 page summary goes here}

\pagebreak

@@

Because progress in scientific research is dependent on the quality and
accessibility of software at all levels, it is now critical to address many
challenges related to the development, deployment, and maintenance of reusable
software.
%\note{Ketan: What are the levels in scientific process where software
%is required. It might be interesting to outline these stages and types of
%software required in each stage.}
In addition, it is essential that scientists,
researchers, and students are able to learn and adopt software-related skills
and methodologies. Established researchers are already acquiring some of these
skills, and in particular a specialized class of software developers is
emerging in academic environments who are an integral and embedded part of
successful research teams. The WSSSPE workshop was intended to provide a forum
for discussion of the challenges, including both positions and experiences. The
short papers and debates have been archived to provide a basis for continued
discussion, and the workshop has fed into the collaborative writing of this
document. An estimate of 90 to 150 participants were present
throughout the day. Additional papers based on extended versions of workshop
submissions are expected. The level of interest in the workshop has led the
organizers, working with some of the submitters and attendees, to plan two
additional workshops (at the 2014 SciPy and SC14 conferences) and to turn the
workshop website into a community website that can be used as a focus for
further discussion and progress. Additionally, a minisymposium that aims to
further explore some of the key issues raised in WSSSPE is co-organized by a
WSSSPE1 participant at the 2014 SIAM Annual Meeting on ``Reliable
Computational Science.''

Before the WSSSPE1 workshop took place, the organizers self-published
a paper~\cite{WSSSPE1-pre-report} to document the process of
organizing and advertising the workshop, collecting and reviewing the
papers, and putting together the agenda. Section~\ref{sec:process} is
based on that publication. The remainder of this paper is based on the
workshop itself, as documented in a set of collaborative notes taken
during the workshop~\cite{WSSSPE1-google-notes}, including discussion
about the keynote presentations (\S\ref{sec:keynotes}), the three
panels (\S\ref{sec:devel}-\ref{sec:community}), 
cross-cutting issues (\S\ref{sec:cross-cutting}),
use cases (\S\ref{sec:use-cases}), and conclusions
(\S\ref{sec:conclusions}).

@@

\begin{quote}
In practice, scientific software activities are part of an ecosystem
where key roles are held by developers, users, and funders. All three
groups supply resources to the ecosystem, as well as requirements that
bound it. Roughly following the example of NSF's Vision and Strategy
for Software~\cite{NSF_software_vision},
the ecosystem may be viewed as having challenges related to:

\begin{itemize}[leftmargin=0.2in]
\item the development process that leads to new (versions of) software
\begin{itemize}[leftmargin=0.2in]
\item how fundamental research in computer science or
  science/engineering domains is turned into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of
    scientific software and vice versa
\end{itemize}
\item the support and maintenance of existing software, including
  software engineering
\begin{itemize}[leftmargin=0.2in]
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and
  sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}[leftmargin=0.2in]
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}[leftmargin=0.2in]
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and multiple countries,
  such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of
  publishers
\end{itemize}
\item education and training
\end{itemize}

@@

\begin{quote}
We invite short (4-page) position/experience reports that will be used
to organize panel and discussion sessions. These papers will be
archived by a third-party service, and provided DOIs. We encourage
submitters to license their papers under a Creative Commons license
that encourages sharing and remixing, as we will combine ideas (with
attribution) into the outcomes of the workshop.  An interactive site
will be created to link these papers and the workshop discussion, with
options for later comments and contributions. Contributions will be
peer-reviewed for relevance and originality before the links are added
to the workshop site; contributions will also be used to determine
discussion topics and panelists. We will also plan one or more papers
to be collaboratively developed by the contributors, based on the
panels and discussions.
\end{quote}

@@

The organizers decided to list 54 of the papers as significant
contributions to the workshop, a very high acceptance rate, but one
that is reasonable, given the goal of broad participation and the fact
that the reports were already self-published. The papers were
grouped into three main categories, namely \emph{Developing and
Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
subject area was associated with a panel and discussion at the
workshop.

@@

\todo{section should include what was presented and what was discussed}

@@

Communities are extremely important in software projects, and both their
building and continued engagement need attention during the project life cycle.
Several of the submitted papers discussed how communities have been built around
projects, and what is needed to enable a project to grow. Among these are public
source code hosting, mailing lists, documentation, wikis, bug trackers, software
downloads, continuous integration, software quality dashboards, and of course,
a general web presence to tie all of these things together. Questions were posed
around users not answering the questions of other users. Several participants
offered counterexamples, such as mailing lists where developers do not participate
as much due to users being more active, or whether the ``core
team'' can end up setting unrealistic expectations by doing too much. Team Geek\todo{add cite}
and Turk's paper on scaling code in the human dimension\todo{add cite}  discuss how development
lists that are welcoming to people actually have many more people contributing.

@@

\subsection*{Papers}

@@

\subsection*{Modeling Sustainability}

\todo{seems like definitions will be mentioned elsewhere in the paper?}
Each of the framework papers included a definition of sustainability, with many overlaps.  In essence the papers related the question to change over time, an idea well captured in the definition used by the UK's Scientific Software Sustainability Institute: ``software you use today will be available---and continue to be improved and supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}. This idea was also expressed as software that continues to serve its users~\cite{Pierce_WSSSPE}.

\subsection*{Credit, Citation, Impact}

\subsection*{Implementing Policy}

Two of the papers in this group focus on specific facets of software
design that are important factors in a project's sustainability but
are often addressed only late in the scientific software development
cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API
governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity
models for software security. The other two papers discuss
implementation strategies for science from the perspective of
facilitating many or all facets of sustainanability-oriented software
design: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
large projects that have software infrastructue development built-in,
with cross-training domain scientist PIs in software engineering best
practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
specialized software engineering skills could be turned into shared
instrumentation with low barriers to access.

API governance and cybersecurity measures appropriate for a project's
maturity are all but two facets of sustainability-oriented software
development. Others include user-centered interface design, test
engineering, dependency management, and deployment operations. Each of
these facets require specialized skills and training in software
engineering. How can the implementation of best practices and
guidelines along many or all of these different facets be facilitated
in scientific software development? Blanton and
Lenhardt~\cite{Blanton_WSSSPE} contrast two models. In one, the time
and personnel devoted to software engineering is ``co-funded'' with
the driving science project. This typically implies large multi-year
collaborative projects that to succeed require significant software
infrastructure to be built, and which thus have the funding to support
one or several software engineers. The sustainability of such projects
then depends on sustaining the funding. In the other model, domain
scientists also take the role of software developers, whether by
necessity such as funding limitations, or due to cross-disciplinary
professional interests. For this to result in sustainable software
products, the domain scientists need to be (or become) cross-trained
in software engineering standards and best practices.

In practice, there are example for both extremes of the spectrum. For
example, in the life sciences the iPlant Collaborative~\cite{iPlant},
the Galaxy Project~\cite{Galaxy}, and Qiime~\cite{Qiime} are large
multi-year projects with significant software infrastructure funding
and deliverables. Typically though these are the exception rather than
the rule, and particularly in the long tail of science the
scientist-developer predominates. For providing these with better
training in software engineering best practices, initiatives such as
Software Carpentry~\cite{SoftwareCarpentry} have been especially
valuable, and Software Carpentry also shown how such training could
scale.
